---
title: "Interactive Plotly Charts"
output:
  html_document:
    theme: flatly
    self_contained: true
    css: !expr NULL   # placeholder — inline CSS below handles styling
---

```{=html}
<style>
/* ── Layout ── */
body { font-family: 'Georgia', serif; margin: 0; background: #f8f9fa; }
h1 { text-align: center; padding: 3rem 1rem 1rem; font-size: 2.2rem; color: #222; }
.intro { max-width: 680px; margin: 0 auto 3rem; font-size: 1.15rem;
         line-height: 1.7; color: #444; text-align: center; padding: 0 1rem; }

/* ── Scrollytelling layout ── */
.scroll-section {
  display: flex;
  gap: 2rem;
  max-width: 1100px;
  margin: 0 auto 4rem;
  padding: 0 1.5rem;
  align-items: flex-start;
  min-height: 80vh;
}
.scroll-narrative {
  flex: 0 0 340px;
  position: sticky;
  top: 80px;
  align-self: flex-start;
}
.scroll-steps { flex: 1; }
.step {
  background: white;
  border-radius: 12px;
  padding: 1.5rem;
  margin-bottom: 2rem;
  box-shadow: 0 2px 12px rgba(0,0,0,0.07);
}
.step h2 { font-size: 1.3rem; color: #1a1a2e; margin-top: 0; }
.step p  { color: #555; line-height: 1.65; font-size: 0.97rem; }

/* ── Sticky narrative card ── */
.narrative-card {
  background: white;
  border-radius: 12px;
  padding: 1.5rem;
  box-shadow: 0 2px 12px rgba(0,0,0,0.07);
  border-left: 4px solid #6c63ff;
  transition: border-color 0.3s;
  min-height: 180px;
}
.narrative-card h3 { margin-top: 0; color: #6c63ff; font-size: 1rem;
                     text-transform: uppercase; letter-spacing: 0.05em; }
.narrative-card p  { font-size: 0.95rem; color: #444; line-height: 1.65; }

/* ── Progress indicator ── */
.progress-bar {
  position: fixed; top: 0; left: 0; height: 4px;
  background: linear-gradient(90deg, #6c63ff, #ff6584);
  width: 0%; z-index: 1000; transition: width 0.1s;
}

/* ── Single-column fallback for mobile ── */
@media (max-width: 768px) {
  .scroll-section { flex-direction: column; }
  .scroll-narrative { position: static; flex: unset; }
}
</style>

<div class="progress-bar" id="progress"></div>
```

```{r setup, include=FALSE}
library(plotly)
library(ggplot2)
library(dplyr)
library(readr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Getting Started with Plotly

::: {.intro}
Plotly makes interactive charts easy in R. Scroll through the sections below —
each chart appears alongside its explanation. Hover over any chart to explore the data.
:::

---

```{=html}
<div class="scroll-section">
  <!-- Sticky narrative (left) -->
  <div class="scroll-narrative">
    <div class="narrative-card" id="narrative-card">
      <h3 id="nav-label">📊 Chart Guide</h3>
      <p id="nav-text">Scroll down to explore different plotly chart types,
         from box plots to dumbbell charts.</p>
    </div>
  </div>

  <!-- Scrolling steps (right) -->
  <div class="scroll-steps">
```

### Box Plots {#box}

```{r}
plot_ly(midwest, x = ~percollege, color = ~state, type = "box", height = 300)
```

**Distribution of college education across Midwest states.** Illinois and Michigan
show the widest spread — some counties are highly educated, others much less so.

---

### Bar Charts {#bar}

```{r}
plot_ly(
  x = c("giraffes", "orangutans", "monkeys"),
  y = c(20, 14, 23),
  name = "SF Zoo",
  type = "bar",
  height = 250,
  marker = list(color = c("#6c63ff", "#ff6584", "#43aa8b"))
)
```

**Categorical comparisons with `type = "bar"`.** The `type` argument controls
chart style — try `"scatter"`, `"histogram"`, or `"heatmap"` for different looks.

---

### Scatter Plots {#scatter}

```{r}
plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length,
        color = ~Species, height = 350,
        type = "scatter", mode = "markers",
        marker = list(size = 8, opacity = 0.75))
```

**Relationships between continuous variables.** Petal length and sepal length
are strongly correlated, and the three *Iris* species naturally separate in space.

---

### ggplot2 → Plotly {#ggplotly}

```{r}
p <- ggplot(mtcars, aes(x = mpg, fill = factor(cyl))) +
  geom_dotplot(stackgroups = TRUE, binwidth = 1, binpositions = "all") +
  scale_fill_brewer(palette = "Set2", name = "Cylinders") +
  labs(x = "Miles per Gallon") +
  theme_minimal()
ggplotly(p, height = 300)
```

**`ggplotly()` wraps any ggplot2 chart instantly.** Build in ggplot2 for
fine-grained control, then hand off to `ggplotly()` for hover interactivity.

---

### Dumbbell Plots {#dumbbell}

```{r}
gapminder <- read_csv(
  "https://raw.githubusercontent.com/datavizpyr/data/master/gapminder-FiveYearData.csv",
  show_col_types = FALSE
)
df <- gapminder |>
  filter(year %in% c(1952, 2007), continent == "Asia") |>
  mutate(paired = rep(1:(n() / 2), each = 2), year = factor(year))

p <- df |> group_by(paired) |>
  ggplot(aes(x = lifeExp, y = reorder(country, lifeExp))) +
  geom_line(aes(group = paired), color = "grey70") +
  geom_point(aes(color = year), size = 2.5) +
  scale_color_viridis_d(name = "Year", option = "turbo") +
  labs(y = NULL, x = "Life Expectancy") +
  theme_minimal(base_size = 10)
ggplotly(p, height = 480)
```

**Change between two time points.** Each line connects 1952 to 2007 for an
Asian country. The dramatic rightward shift reflects one of humanity's greatest
public health achievements.

---

```{=html}
  </div><!-- scroll-steps -->
</div><!-- scroll-section -->
```

```{=html}
<script>
// Update sticky narrative card as user scrolls
const steps = [
  { id: "box",       label: "📦 Box Plots",
    text: "Box plots show the spread and skew of a distribution. The IQR (box) contains the middle 50% of data." },
  { id: "bar",       label: "📊 Bar Charts",
    text: "Use type = 'bar' for categorical comparisons. Plotly supports bar, scatter, box, heatmap, and more." },
  { id: "scatter",   label: "🔵 Scatter Plots",
    text: "type = 'scatter' with mode = 'markers' gives you a classic scatter. Add color to encode a third variable." },
  { id: "ggplotly",  label: "🔄 ggplot2 → Plotly",
    text: "ggplotly() converts any ggplot2 object. Great for reusing your existing ggplot2 code with hover." },
  { id: "dumbbell",  label: "🏋️ Dumbbell Plots",
    text: "Dumbbells highlight change between two states. Connect two points per item to show before/after." },
];

const navLabel = document.getElementById("nav-label");
const navText  = document.getElementById("nav-text");
const progress = document.getElementById("progress");
const card     = document.getElementById("narrative-card");

const colors   = ["#6c63ff","#ff6584","#43aa8b","#f7a325","#e63946"];

function updateNarrative(idx) {
  const s = steps[idx];
  navLabel.textContent = s.label;
  navText.textContent  = s.text;
  card.style.borderLeftColor = colors[idx];
  navLabel.style.color = colors[idx];
}

function onScroll() {
  // Progress bar
  const scrolled = window.scrollY;
  const total    = document.body.scrollHeight - window.innerHeight;
  progress.style.width = (scrolled / total * 100) + "%";

  // Active step
  let active = 0;
  steps.forEach((s, i) => {
    const el = document.getElementById(s.id);
    if (!el) return;
    const rect = el.getBoundingClientRect();
    if (rect.top < window.innerHeight * 0.55) active = i;
  });
  updateNarrative(active);
}

window.addEventListener("scroll", onScroll, { passive: true });
onScroll();
</script>
```
