---
title: "Interactive Charts with ggiraph"
format:
  closeread-html:
    cr-style:
      section-background-color: "#f8f9fa"
      narrative-background-color: "#ffffff"
      narrative-font-size: 1.1em
      sticky-col-width: 60%
---

```{r setup, include=FALSE}
library(ggiraph)
library(ggplot2)
library(dplyr)
library(readr)
```

# Getting Started with ggiraph {.unnumbered}

ggiraph extends ggplot2 with interactivity — hover tooltips, highlights, and click actions —
using the same grammar you already know. As you scroll, each chart will appear alongside explanations.

---

:::{.cr-section}

## Box Plots {.unnumbered}

:::{#cr-boxplot}
```{r}
p <- ggplot(midwest, aes(x = state, y = percollege, fill = state)) +
  geom_boxplot(alpha = 0.7) +
  geom_point_interactive(
    aes(tooltip = paste0(state, "\nCounty % college: ", round(percollege, 1), "%"),
        data_id = state),
    position = position_jitter(width = 0.2, seed = 42),
    size = 1.5, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = NULL, y = "% College Educated",
       title = "College Education by Midwest State") +
  theme_minimal(base_size = 13)

girafe(ggobj = p, width_svg = 6, height_svg = 4,
       options = list(opts_hover(css = "opacity:1;stroke:black;stroke-width:2px;"),
                      opts_tooltip(css = "background:white;padding:6px;border-radius:4px;
                                          border:1px solid #ccc;font-size:13px;")))
```
:::

Box plots are great for comparing distributions across groups.
Here we're looking at the **percentage of college-educated residents**
across Midwestern states. **Hover over each box** to see the state name and value.
Notice how Illinois has a wider spread compared to Indiana.

[@cr-boxplot]{scale-by="0.9"}

---

## Bar Charts {.unnumbered}

:::{#cr-bar}
```{r}
zoo_df <- data.frame(
  animal = c("giraffes", "orangutans", "monkeys"),
  count  = c(20, 14, 23)
)

p <- ggplot(zoo_df, aes(x = animal, y = count, fill = animal,
            tooltip = paste0(animal, ": ", count),
            data_id = animal)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("#2196F3", "#4CAF50", "#FF9800"), guide = "none") +
  labs(title = "SF Zoo Animal Count", x = NULL, y = "Count") +
  theme_minimal(base_size = 13)

p
# girafe(ggobj = p, width_svg = 6, height_svg = 3.5,
#        options = list(opts_hover(css = "opacity:1;filter:brightness(1.15);"),
#                       opts_tooltip(css = "background:white;padding:6px;border-radius:4px;
#                                           border:1px solid #ccc;font-size:13px;")))
```
:::

Bar charts are ideal for **categorical comparisons**.
With ggiraph, swap `geom_col()` for `geom_col_interactive()` and add a
`tooltip` aesthetic — that's all it takes. **Hover over each bar** to see the value.

[@cr-bar]

---

## Scatter Plots {.unnumbered}

:::{#cr-scatter}
```{r}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species,
            tooltip = paste0("Species: ", Species,
                             "\nSepal: ", Sepal.Length,
                             "\nPetal: ", Petal.Length),
            data_id = paste0(Species, Sepal.Length, Petal.Length))) +
  geom_point_interactive(size = 2.5, alpha = 0.75) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Iris: Sepal vs Petal Length",
       x = "Sepal Length", y = "Petal Length") +
  theme_minimal(base_size = 13)

girafe(ggobj = p, width_svg = 6, height_svg = 4,
       options = list(opts_hover(css = "r:5px;opacity:1;stroke:black;stroke-width:1px;"),
                      opts_tooltip(css = "background:white;padding:6px;border-radius:4px;
                                          border:1px solid #ccc;font-size:13px;")))
```
:::

Scatter plots reveal **relationships between two continuous variables**.
The iris dataset shows a clear positive correlation between sepal length
and petal length. **Hover over any point** to see the exact measurements.
The three species separate naturally in space.

[@cr-scatter]{scale-by="1.1"}

---

## Dot Plots {.unnumbered}

:::{#cr-dotplot}
```{r}
p <- ggplot(mtcars, aes(x = mpg, fill = factor(cyl),
            tooltip = paste0("MPG: ", mpg, "\nCylinders: ", cyl),
            data_id = row.names(mtcars))) +
  geom_dotplot_interactive(stackgroups = TRUE, binwidth = 1,
                           binpositions = "all", alpha = 0.85) +
  scale_fill_brewer(palette = "Set2", name = "Cylinders") +
  labs(title = "MPG Distribution by Cylinder Count", x = "Miles per Gallon") +
  theme_minimal(base_size = 13)

girafe(ggobj = p, width_svg = 6, height_svg = 3.5,
       options = list(opts_hover(css = "opacity:1;stroke:black;stroke-width:1.5px;"),
                      opts_tooltip(css = "background:white;padding:6px;border-radius:4px;
                                          border:1px solid #ccc;font-size:13px;")))
```
:::

With ggiraph, the pattern is always the same: take any `geom_*()` from ggplot2,
add `_interactive`, and pass `tooltip` and `data_id` aesthetics.
**Hover over the dots** to see each car's MPG and cylinder count.

[@cr-dotplot]

---

## Dumbbell Plots {.unnumbered}

:::{#cr-dumbbell}
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("gapminder.csv", show_col_types = FALSE)

df <- gapminder |>
  filter(year %in% c(1952, 2007), continent == "Asia") |>
  mutate(year = factor(year))

p <- ggplot(df, aes(x = lifeExp, y = reorder(country, lifeExp))) +
  geom_line(aes(group = country), color = "grey70", linewidth = 0.8) +
  geom_point_interactive(
    aes(color = year,
        tooltip = paste0(country, " (", year, ")\nLife Exp: ", round(lifeExp, 1)),
        data_id = paste0(country, year)),
    size = 3) +
  scale_color_viridis_d(name = "Year", option = "turbo") +
  labs(y = NULL, x = "Life Expectancy",
       title = "Life Expectancy in Asia: 1952 vs 2007") +
  theme_minimal(base_size = 10)

p
# 
# girafe(ggobj = p, width_svg = 6, height_svg = 7,
#        options = list(opts_hover(css = "r:6px;opacity:1;stroke:black;stroke-width:1px;"),
#                       opts_tooltip(css = "background:white;padding:6px;border-radius:4px;
#                                           border:1px solid #ccc;font-size:13px;")))
```
:::

Dumbbell plots excel at showing **change between two time points**.
Each line connects 1952 (purple) to 2007 (yellow) for each Asian country.
**Hover over any point** to see the exact life expectancy.
The dramatic rightward shift represents one of the greatest improvements
in human longevity in recorded history.

[@cr-dumbbell]{scale-by="0.95"}

:::
