---
title: "Interactive Plotly Charts"
format:
  closeread-html:
    cr-style:
      section-background-color: "#f8f9fa"
      narrative-background-color: "#ffffff"
      narrative-font-size: 1.1em
      sticky-col-width: 60%
---

```{r setup, include=FALSE}
library(plotly)
library(ggplot2)
library(dplyr)
library(readr)
```

# Getting Started with Plotly {.unnumbered}

Plotly makes it easy to create beautiful interactive visualizations in R.
As you scroll, each chart will appear in sequence alongside explanations.

---

:::{.cr-section}

## Box Plots {.unnumbered}

:::{#cr-boxplot}
```{r}
fig <- plot_ly(midwest, x = ~percollege, color = ~state, type = "box", height = 400)
fig
```
:::

Box plots are great for comparing distributions across groups.
Here we're looking at the **percentage of college-educated residents**
across Midwestern states. Notice how Illinois and Michigan have wider
spreads compared to Indiana.

[@cr-boxplot]{scale-by="0.9"}

---

## Bar Charts {.unnumbered}

:::{#cr-bar}
```{r}
fig <- plot_ly(
  x = c("giraffes", "orangutans", "monkeys"),
  y = c(20, 14, 23),
  name = "SF Zoo",
  type = "bar",
  height = 350,
  marker = list(color = c("#2196F3", "#4CAF50", "#FF9800"))
)
fig
```
:::

Bar charts are ideal for **categorical comparisons**. 
The `type = "bar"` argument is the key — plotly supports many chart types 
(scatter, box, histogram, heatmap) using the same syntax.

[@cr-bar]

---

## Scatter Plots {.unnumbered}

:::{#cr-scatter}
```{r}
plot_ly(data = iris,
        x = ~Sepal.Length,
        y = ~Petal.Length,
        color = ~Species,
        height = 400,
        type = "scatter",
        mode = "markers",
        marker = list(size = 8, opacity = 0.7))
```
:::

Scatter plots reveal **relationships between two continuous variables**.
The iris dataset shows a clear positive correlation between sepal length
and petal length — and the species separate naturally in space.

[@cr-scatter]{scale-by="1.1"}

---

## Converting ggplot2 → Plotly {.unnumbered}

:::{#cr-ggplotly}
```{r message=FALSE, warning=FALSE}
p <- ggplot(mtcars, aes(x = mpg, fill = factor(cyl))) +
  geom_dotplot(stackgroups = TRUE, binwidth = 1, binpositions = "all") +
  scale_fill_brewer(palette = "Set2", name = "Cylinders") +
  labs(title = "MPG Distribution by Cylinder Count", x = "Miles per Gallon") +
  theme_minimal()

ggplotly(p, height = 350)
```
:::

The `ggplotly()` function wraps any ggplot2 object into an interactive plot.
This is the fastest route to interactivity — build your chart in ggplot2,
then simply pass it to `ggplotly()`. Hover over the dots to see exact values!

[@cr-ggplotly]

---

## Dumbbell Plots {.unnumbered}

:::{#cr-dumbbell}
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv(
  "https://raw.githubusercontent.com/datavizpyr/data/master/gapminder-FiveYearData.csv",
  show_col_types = FALSE
)

df <- gapminder |>
  filter(year %in% c(1952, 2007), continent == "Asia") |>
  mutate(paired = rep(1:(n() / 2), each = 2),
         year = factor(year))

p <- df |>
  group_by(paired) |>
  ggplot(aes(x = lifeExp, y = reorder(country, lifeExp))) +
  geom_line(aes(group = paired), color = "grey70") +
  geom_point(aes(color = year), size = 2.5) +
  scale_color_viridis_d(name = "Year", option = "turbo") +
  labs(y = NULL, x = "Life Expectancy",
       title = "Life Expectancy in Asia: 1952 vs 2007") +
  theme_minimal(base_size = 10)

ggplotly(p, height = 500)
```
:::

Dumbbell plots excel at showing **change between two time points**.
Each line connects 1952 (purple) to 2007 (yellow) for each Asian country.
The dramatic rightward shift represents one of the greatest improvements
in human longevity in recorded history.

[@cr-dumbbell]{scale-by="0.95"}

:::
